\section{The grammar of Plutus Core}
\label{sec:untyped-plc-grammar}
This section presents the grammar of Plutus Core in a Lisp-like form.  This is
intended as a specification of the abstract syntax of the language; it may also
by used by tools as a concrete syntax for working with Plutus Core programs, but
this is a secondary use and we do not make any guarantees of its completeness
when used in this way.  The primary concrete form of Plutus Core programs is the
binary format described in Appendix~\ref{appendix:flat-serialisation}.

\subsection{Lexical grammar}
\label{sec:untyped-plc}
\thispagestyle{plain}
\pagestyle{plain}

\begin{minipage}{\linewidth}
    \centering
    \[\begin{array}{lrclr}

        \textrm{Name}        & n      & ::= & \texttt{[a-zA-Z][a-zA-Z0-9\_\textquotesingle]\textsuperscript{*}}   & \textrm{name}\\

        \textrm{Var}           & x      & ::= & n & \textrm{term variable}\\
        \textrm{BuiltinName}   & bn     & ::= & n & \textrm{built-in function name}\\
        \textrm{Version} & v & ::= & \texttt{[0-9]\textsuperscript{+}.[0-9]\textsuperscript{+}.[0-9]\textsuperscript{+}}& \textrm{version}\\

        \textrm{Constant} & c & ::= & \langle{\textrm{literal constant}}\rangle& \\
        \textrm{Natural}  & n      & ::= & [0-9]+ & \textrm{A natural}\\

    \end{array}\]
    \captionof{figure}{Lexical grammar of Plutus Core}
    \label{fig:lexical-grammar-untyped}
\end{minipage}%
\nomenclature[C]{$L,M,N$}{A term}%
\nomenclature[C]{$n$}{A name}%
\nomenclature[C]{$n$}{An natural}%
\nomenclature[C]{$x$}{A variable name}%
\nomenclature[C]{$bn, b$}{The name of a built-in function}%
\nomenclature[C]{$c$}{A literal constant}%
\nomenclature[C]{$P$}{A Plutus Core program}%
\nomenclature[C]{$v$}{Plutus Core version}



%   @sqs   = '  ( ($printable # ['\\])  | (\\$printable) )* '
%   
%   -- A double quoted string, allowing escaped characters including \".  Similar to @sqs
%   @dqs   = \" ( ($printable # [\"\\]) | (\\$printable) )* \"
%   
%   -- A sequence of printable characters not containing '(' or ')' such that the
%   -- first character is not a space or a single or double quote.  If there are any
%   -- further characters then they must comprise a sequence of printable characters
%   -- possibly including spaces, followed by a non-space character.  If there are
%   -- any leading or trailing spaces they will be consumed by the $white+ token
%   -- below.
%   $nonparen = $printable # [\(\)]
%   @chars = ($nonparen # ['\"$white]) ($nonparen* ($nonparen # $white))?
%   
%       <literalconst> "()" | @sqs | @dqs | @chars { tok (\p s -> alex $ TkLiteralConst p (textOf s)) `andBegin` 0 }

%% "()"
%% @sqs   = '  ( ($printable # ['\\])  | (\\$printable) )* '
%% @dqs   = \" ( ($printable # [\"\\]) | (\\$printable) )* \"
%% @chars = ($nonparen # ['\"$white]) ($nonparen* ($nonparen # $white))?


\subsection{Grammar}
\begin{minipage}{\linewidth}
    \centering
    \[\begin{array}{lrclr}
    \textrm{Term}       & L,M,N  & ::= & x                               & \textrm{variable}\\
                        &        &     & \con{\tn}{c}                    & \textrm{constant}\\
                        &        &     & \builtin{b}                     & \textrm{builtin}\\
                        &        &     & \lamU{x}{M}                     & \textrm{$\lambda$ abstraction}\\
                        &        &     & \appU{M}{N}                     & \textrm{function application}\\
                        &        &     & \delay{M}                       & \textrm{delay execution of a term}\\
                        &        &     & \force{M}                       & \textrm{force execution of a term}\\
                        &        &     & \constr{n}{M_0 \ldots M_{m-1}}  & \textrm{constructor with tag $n$ and $m$ arguments}\\
                        &        &     & \kase{M}{N_0 \ldots N_{m-1}}    & \textrm{case analysis with $m$ alternatives}\\
                        &        &     & \errorU                         & \textrm{error}\\
        \textrm{Program}& P      & ::= & \version{v}{M}                  & \textrm{versioned program}

    \end{array}\]
    \captionof{figure}{Grammar of untyped Plutus Core}
    \label{fig:untyped-grammar}
\end{minipage}


\subsection{Notes}
\label{sec:grammar-notes}
\paragraph{Version numbers.} The version number at the start of a program specifies
the Plutus Core language version used in the program.

A \textit{Plutus Core language version} describes a version of
the basic language with a particular set of features. A language version
consists of three non-negative integers separated by decimal points, for
example \texttt{1.4.2}. Language versions are ordered lexicographically.

The grammar above describes Plutus Core version 1.1.0. Version 1.0.0
is identical, except that \texttt{constr}
and \texttt{case} are not included.  Version 1.0.0 is fully forward-compatible
with version 1.1.0, so any valid version 1.0.0 program is also a valid version
1.1.0 program.  The semantics, evaluator and serialisation formats described
later in this document all apply to both versions, except that it is an error to
use \texttt{constr} or \texttt{case} in any program with a version prior to
1.1.0: a parser, deserialiser, or evaluator should fail immediately if
\texttt{constr} or \texttt{case} is encountered when processing such a program.

\paragraph{Scoping.} For simplicity, \textbf{we assume throughout that the body of a
Plutus Core program is a closed term}, ie, that it contains no free variables.
Thus \texttt{(program 1.0.0 (lam x x))} is a valid program but \texttt{(program
  1.0.0 (lam x y))} is not, since the variable \texttt{y} is free. This
condition should be checked before execution of any program commences, and the
program should be rejected if its body is not closed.  The assumption implies
that any variable $x$ occurring in the body of a program must be bound by an
occurrence of \texttt{lam} in some enclosing term; in this case, we always
assume that $x$ refers to the \textit{most recent} (ie, innermost) such binding.

\paragraph{Iterated applications.}
An application of a term $M$ to a term $N$ is represented by
$\appU{M}{N}$. We may occasionally write
$\appU{M}{N_1 \ldots N_k}$ or
$\appU{M}{\repetition{N}}$ as an abbreviation for an iterated application
$\mathtt{[}\ldots\mathtt{[[}M\;N_1\mathtt{]}\;N_2\mathtt{]}\ldots $  %% Try to avoid a bad line break
  $N_k\mathtt{]}$,
and tools may also use this as concrete syntax.

\paragraph{Constructors and case analysis}
Plutus Core supports creating structured data using $\keyword{constr}$ and deconstructing
it using $\keyword{case}$. Both of these terms are unusual in that they have (possibly
empty) lists of children: $\keyword{constr}$ has the (0-based) \emph{tag} and then a list of arguments;
$\keyword{case}$ has a scrutinee and then a list of case branches. Their behaviour is mostly
straightforward: $\keyword{constr}$ evaluates its arguments and forms a value; $\keyword{case}$
evaluates the scrutinee into a $\keyword{constr}$ value, selects the branch corresponding to
the tag on the value, and then applies that to the arguments in the value. The only thing
to note is that $\keyword{case}$ does \emph{not} strictly evaluate the case branches, only
applying (and hence evaluating) the one that is eventually selected.

\paragraph{Constructor tags}
Constructor tags can in principle be any natural number. In practice, since they cannot
be dynamically constructed, we can limit them to a fixed size without having to worry
about overflow. So we limit them to 64 bits, although this is currently only enforced in
the binary format (Appendix~\ref{appendix:flat-serialisation}).

\paragraph{Built-in types and functions.} The language is parameterised by a set $\Uni$ of
\textit{built-in types} (we sometimes refer to $\Uni$ as the \textit{universe})
and a set $\Fun$ of \textit{built-in functions} (\textit{builtins} for short),
both of which are sets of Names.  Briefly, the built-in types represent sets of
constants such as integers or strings; constant expressions $\con{\tn}{c}$
represent values of the built-in types (the integer 123 or the string
\texttt{"string"}, for example), and built-in functions are functions operating
on these values, and possibly also general Plutus Core terms.  Precise details
are given in Section~\ref{sec:specify-builtins}.

See Section~\ref{sec:cardano-builtins}
for a description of the types and functions which have already been deployed on
the Cardano blockchain (or will be in the near future).

\nomenclature[E1]{$\Fun$}{The set of built-in functions}

\paragraph{De Bruijn indices.}
The grammar defines names to be textual strings, but occasionally (specifically
in Appendix~\ref{appendix:flat-serialisation}) we want to use de Bruijn indices
(\cite{deBruijn}, \cite[C.3]{Barendregt}), and for this we redefine names to be
natural numbers.  In de Bruijn terms, $\lambda$-expressions do not need to bind
a variable, but in order to re-use our existing syntax we arbitrarily use 0 for
the bound variable, so that all $\lambda$-expresssions are of the form
\texttt{(lam 0 $M$)}; other variables (ie, those not appearing immediately after
a \texttt{lam} binder) are represented by natural number greater than zero.

\paragraph{Lists in constructor and case terms}
The grammar defines constructor and case terms to have a variable number of
subterms written in sequence with no delimiters. This corresponds to the
concrete syntax, e.g. we write $\constr{0}{t_1\ t_2\ t_3}$. However, in the
rest of the specification we will abuse notation and treat these terms as
having \emph{lists} of subterms.
