\section{Version numbers and the Cardano blockchain}
\label{appendix:version-numbers}
After a Plutus Core script has been deployed to the blockchain it may be
re-evaluated at any point in the future when the history of the chain is
verified.  Plutus Core may evolve as time progresses, and because of this we
have to be quite careful with versioning: a detailed discussion of this issue
can be found in~\cite{CIP-35}.  There are two types of version which are
relevant in this document:

\begin{itemize}
\item The language itself may evolve: for example some fundamental new construct
  may be added.  A \textit{Plutus Core language version} describes a version of
  the basic language with a particular set of features. A language version
  consists of three non-negative integers separated by decimal points, for
  example \texttt{1.4.2}.  Language versions are ordered lexicographically.  As
  we have seen in Section~\ref{sec:grammar-notes}, the Plutus Core
  language version used by a script is included in the \texttt{program} header.
  
\item The built-in types and functions deployed on the chain along with a
  particular version of the Plutus Core language can change with time.  Plutus
  Core scripts on the Cardano chain are equipped with a tag called a
  \textit{ledger language version} which (among other things) determines which
  set of types and functions are to be used during evaluation of the script.
  Ledger language versions have names names like ``PlutusV1'', ``PlutusV2'',
  and so on.  The semantics of a given built-in function may be different in
  different ledger language versions.  Each Plutus Core ledger language version
  is only valid with a specific Plutus Core language version, but a given
  language version may support multiple ledger language versions.
\end{itemize}

\subsection{Cardano support for different Plutus Core language and ledger language versions}

Cardano introduced support for Plutus Core in the Alonzo release;
Table~\ref{table:versions} shows all releases that have introduced new language
versions and/or new ledger language versions since then.  The features provided
by the different ledger langauge versions are described in subsequent appendices.
\begin{table}[H]
  \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Cardano release & Date & Language versions & Ledger language versions \\
        \hline
        Alonzo & September 2021 & 1.0.0 & PlutusV1 \\
        Vasil & June 2022 & 1.0.0 & PlutusV1, PlutusV2 \\
        (Forthcoming) & TBA & 1.0.0, 1.1.0 & PlutusV1, PlutusV2, PlutusV3 \\
        \hline
    \end{tabular}
    \caption{Plutus Core language versions and ledger language versions}
    \label{table:versions}
\end{table}
   

\noindent Table~\ref{table:version-dependencies} shows which ledger language
versions are compatible with each Plutus Core language version: if an on-chain
script specifies a combination which does not appear in this table then
attempting to evaluate it will cause an error.
 
%% See PlutusLedgerApi.Common.Versions.plcVersionsIntroducedIn
\begin{table}[H]
  \centering
    \begin{tabular}{|l|l|}
        \hline
        \thead{Plutus Core \\ language version} & \thead{Compatible ledger \\ language versions}\\
        \hline
        1.0.0 & PlutusV1, PlutusV2\\
        1.0.1 & PlutusV3 \\
        \hline
    \end{tabular}
    \caption{Compatibility of Plutus Core language versions and ledger language versions}
    \label{table:version-dependencies}
\end{table}

