\documentclass[../plutus-core-specification.tex]{subfiles}

\begin{document}

We illustrate the use of Plutus Core by constructing a simple
validator program for the situation described in the previous
section. We present components of this program in a high-level style:
for instance, we write
\begin{lstlisting}
   |\one| : |\unit|
   |\one| = (abs $a$ (type) (lam $x$ $a$ $x$))
\end{lstlisting}
for the element of the type $unit$ (defined in Figure~\ref{fig:type_abbreviations}).

We stress that declarations in this style are not part of the Plutus
Core language. We merely use the familiar syntax to present our
example. If the high-level definitions in our example were compiled to
a Plutus Core expression, it would result in something like
Figure~\ref{fig:Continuized_Let_Example}.

We proceed by defining the booleans. Like \textit{unit}, the type \textit{boolean}
below is an abbreviation in the specification. Some built-in constants
return values of type $boolean$. When needed, user programs should
contain the declarations below.

% Note that |\boolean| below causes \boolean to be treated as normal latex.
% It works like this because of the \escapeinside setting in \lstset in the document header.

\noindent We have

\begin{lstlisting}
  |\true| : |\boolean|
  |\true| = (abs $a$ (type)
          (lam $x$ $a$
          (lam $y$ $a$
                $x$ )))
\end{lstlisting}
and similarly
\begin{lstlisting}
  |\false| : |\boolean|
  |\false| = (abs $a$ (type)
           (lam $x$ $a$
           (lam $y$ $a$
                 $y$ )))
\end{lstlisting}

\noindent Next, we define the \case{} function for the type $boolean$ as follows:
\begin{lstlisting}
  |\case| : (all $a$ (type)
          (fun |\boolean|
          (fun $a$ (fun $a$ $a$))))
  |\case| = (abs $a$ (type)
          (lam $b$ |\boolean|
          (lam $t$ $a$
          (lam $f$ $a$
            [ {$b$ $a$} $t$ $f$ ]
          ))))
\end{lstlisting}
The reader is encouraged to verify that
\begin{lstlisting}
  [{|\case| a} |\true| x y ] $\stackrel{*}{\rightarrow}$ x
\end{lstlisting}
and
\begin{lstlisting}
  [{|\case| a} |\false| x y ] $\stackrel{*}{\rightarrow}$ y
\end{lstlisting}

\noindent We can use \case{} to define the following function:
\begin{lstlisting}
  verifyIdentity :
    (fun (con bytestring) (fun (con bytestring) unit))
  verifyIdentity =
    (lam |\pubkey| (con bytestring)
    (lam |\signed| (con bytestring)
  [
    { |\case| (fun |\unit| |\unit|) }
    [ (builtin verifySignature) |\signed| |\txhash| |\pubkey| ]
    (lam u |\unit| |\one|)
    (lam u |\unit| (error |\unit|))
    |\one|
  ]))
\end{lstlisting}
 Note that although we are producing a value of
type \unit{} we instantiate \case{} at the type \texttt{(fun \unit \unit)} and
pass the branches as functions from \unit{}. This
is because Plutus Core is eagerly evaluated, so if we did not delay the
branches of the case in this way, then the \texttt{(error)} in the second branch
would be evaluated unconditionally, which is not what we want. This way the
branch of the case is selected first, and then the body of the branch is
evaluated.

The idea is that first argument of \texttt{verifyIdentity} is a public
key, and the second argument is the result of signing the hash of the
current transaction (contained in
$\mathit{txhash} : \texttt{(con bytestring)}$) with the corresponding
private key.  The function terminates normally if and only if the
signature is valid, returning \texttt{error} otherwise. Now, given
Alice's public key we can apply our function to obtain one that
verifies whether or not its input is the result of Alice signing the
current block number. Again, we stress that the Plutus Core expression
corresponding to \texttt{verifyIdentity} is going to look something
like Figure \ref{fig:Continuized_Let_Example}.
\todompj{This is out of date. Not sure how useful it is.}

\newpage
\begin{figure*}[h]  % Using H here causes undefined references to this figure
\begin{lstlisting}
(lam |\pubkey| (con bytestring)
(lam |\signed| (con bytestring)
    [
    {
        (abs $a$ (type)
        (lam $b$ (all $a$ (type) (fun $a$ (fun $a$ $a$)))
            (lam $t$ (fun (all $a$ (type) (fun $a$ $a$)) $a$)
            (lam $f$ (fun (all $a$ (type) (fun $a$ $a$)) $a$)
                [
                [ { $b$ (fun (all $a$ (type) (fun $a$ $a$)) $a$) } $t$ $f$ ]
                (abs $a$ (type) (lam $x$ $a$ $x$))
                ]
            )
            )
        )
        )
        (all $a$ (type) (fun $a$ $a$))
    }
    [ (builtin verifySignature) |\signed| |\txhash| |\pubkey| ]
    (lam $u$ (all $a$ (type) (fun $a$ $a$))
        (abs $a$ (type) (lam $x$ $a$ $x$))
    )
    (lam $u$ (all $a$ (type) (fun $a$ $a$))
        (error (all $a$ (type) (fun $a$ $a$)))
    )
    ]
)
)
\end{lstlisting}
\caption{Example of Section 5 written out in full}
\label{fig:Continuized_Let_Example}
\end{figure*}

\end{document}
