\chapter{Plutus Core on Cardano}
\input{cardano/versions.tex}
\section{Built-in types and functions}
\label{sec:cardano-builtins}
\paragraph{Built-in batches.}
\label{sec:builtin-batches}

The built-in types and functions are defined in batches corresponding to how
they were added to ledger languages.  These batches are given in the following
sections.

\paragraph{Built-in semantics variants.}
\label{sec:builtin-semantics-variants}

In rare cases we can make a mistake or need to change the actual behaviour of a
built-in function.  To handle this we define a series of built-in semantics
variants, which indicate which behaviour should be used.  A fix will typically
be deployed by defining a new semantics variant, and then using that variant for
future ledger languages (but not existing ones, since this is usually a
backwards-incompatible change).

Changes are listed alongside the original definition of the built-in function in
its original batch, and are indexed in the following table.

\begin{table}[H]
  \centering
    \begin{tabular}{|l|l|l|}
        \hline
        \thead{Built-in semantics variant} & \thead{Changes from previous semantics} \\
        \hline
        Built-in semantics 1 & None \\
        Built-in semantics 2 & \TT{consByteString} (See~\ref{note:consbytestring}) \\
        \hline
    \end{tabular}
    \caption{Built-in semantics variants}
    \label{table:bs-variants}
\end{table}


\paragraph{Concrete syntax for built-in types.}
Recall that in the abstract notation for built-in types introduced in
Section~\ref{sec:built-in-types}, a built-in type is either an \textit{atomic
  type} such as \texttt{integer} or \texttt{string} or an application $\op(T_1,
\ldots, T_n)$ of a \textit{type operator} to a sequence of built-in types.  The
concrete syntax of built-in types used in textual Plutus Core programs is
slightly different in that we use a curried form of application for type
operators: a type is given by

\newcommand{\bitn}{\mathbf{T}} % built-in type name
\newcommand{\bitc}{\mathbf{C}} % built-in constant syntax

\begin{minipage}{\linewidth}
    \centering
    \[\begin{array}{rclr}
    \bitn & ::= & \textit{atomic-type} & \textrm{Atomic type}\\
                & & \texttt{(}\op \ \bitn_1 \ldots \bitn_{\valency{\op}}) & \textrm{Type application}\\
    \end{array}\]
    \label{fig:built-in-type-concrete-syntax}
\end{minipage}
\noindent Note that we again require that all type operators are fully applied.
We refer to the syntactic objects $\bitn$ above as \textit{concrete built-in
 types}. There is an obvious bijection between these and the abstract built-in
types used elsewhere in this document, and given an abstract built-in type $T$
we will denote the corresponding concrete built-in type by $\bar{T}$.

\paragraph{Concrete syntax for built-in constants.}
We provide concrete syntax for constants of most (but not all) built-in types.
For a built-in type $T$ which has a concrete syntax we specify a set $\bitc_T$
of strings (using either regular expressions or a BNF-style grammar), and a
constant of type $T$ is then represented in the concrete syntax by an expression
of the form \texttt{(con $\bar{T}$ $c_T$)} with $c_T \in \bitc_T$.  Each string
$c_T$ will have an interpretation as a value of type $T$ (ie, an element of
$\denote{T}$) and since this will generally be the obvious interpretation we
will not always spell out the details.%%
\nomenclature[H]{$\bitc_T$}{Set of strings used for the concrete syntax of constants of built-in type $T$}


\input{cardano/builtins1.tex}
\input{cardano/builtins2.tex}
\input{cardano/builtins3.tex}
\input{cardano/builtins4.tex}
\newpage %% Bad page break between table and caption otherwise.
\input{cardano/builtins5.tex}
\input{cardano/builtins6.tex}
\input{cardano/builtins7.tex}


